---
title: "Downloading and using data from bdl"
author: "Krzysztof Kania"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading and using data from bdl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(bdl)
options(tibble.print_min = 4L, tibble.print_max = 4L)
```

The `bdl` package is an interface to Local Data Bank(Bank Danych Lokalnych - bdl) [API](https://bdl.stat.gov.pl/api/v1/home) with set of useful tools like quick 
plotting using data from data bank.

## Intro

Working with `bdl` is based on id codes. Most of data downloading functions requires specifying one or vector of multiple unit or variable ids as a string.

It is recommended to use private API key which u can get [here](https://bdl.stat.gov.pl/api/v1/client).
To apply use: `options(bdl.api_private_key ="your_key")`

Also every function returns data by default in Polish. If you would like get data in English just add `lang = "en"` to any function.

Any metadata information (unit levels, aggregates, NUTS code explanation etc.) can be found [here](https://bdl.stat.gov.pl/BDL/metadane).


## Searching unit id

When searching for unit id we can use two methods:

- Direct search `search_units()`
- Tree listing by `get_units()`

Units consist of 6 levels:
```{r  }
get_levels()
```

Lowest - seventh level has own separate functions with suffix `localities`. Warning - localities functions have different set of arguments. Check package or API documentation for more info.

### Direct search

Direct searching `search_units()` takes couple different arguments like:

- `name` - required search phrase (can be empty string)
- `level` - narrows returned units to given level

and more. To look for more arguments on any given function check package or API documentation.

```{r  }
search_units(name = "wro")
search_units(name = "", level = 3)
```


### Tree listing
To get all units available in local data bank run `get_units()` without any argument(warning - it can eat data limit very fast around 4.5k rows):

To narrow the list add `unitParentId`. Function will return all children units for a given parent at all levels. Add `level` argument to filter units even further.

```{r  }
get_units(parentId = "000000000000", level = 5)
```


## Searching subject and variable id

Subjects are themed directories of variables. 

We have two searching methods for both subjects and variables:

- Direct search `search_variables()` and `search_subjects()`
- Subject tree listing by `get_subjects()` and `get_variables()`


### Subjects

To directly search for subject we just provide search phrase:

```{r  }
search_subjects("lud")
```

Subjects consist of 3 levels (categories, groups, subgroups) - `K`, `G` and `P` respectively. The fourth level of subject (child of subgroup) would be variables.

To list all top level subjects use `get_subjects()`:
```{r  }
get_subjects()
```

To list sub-subjects to given category or group use `get_subjects()` with `parentId` argument:
```{r  }
get_subjects(parentId = "K3")
get_subjects(parentId = "G7")
```


### Variables

Firstly you can list variables for given subject (subgroup):
```{r  }
get_variables("P2425")
```

Secondly you can direct search variables with `search_variables()`. You can use empty string as `name` to list all variables but I strongly advise against as it has around 40 000 rows and you will probably hit data limit.
```{r  }
search_variables("samochod")
```
You can narrow search to given subject - subgroup:
```{r  }
search_variables("lud", subjectId = "P2425")
```

## Downloading data
If you picked unit and variable codes you are ready to download data. You can do this two ways:
- Download data of multiple variables on single unit `get_data_by_unit()`
- Download data of single variable on multiple units `get_data_by_variable()`


### Single unit, multiple variables

We will use `get_data_by_unit()`. We specify our single unit as `unitId` string argument and variables by vector of strings. Optionally we can specify interested years of data. If not all available years.

```{r  }
get_data_by_unit(unitId = "023200000000", varId =  "3643")
get_data_by_unit(unitId = "023200000000", varId =  c("3643", "2137", "148190"))
```

To get more information about data we can add `type` argument and set it to `"label"` to add additional column with variable info.

```{r  }
df <- get_data_by_unit(unitId = "023200000000", varId =  "3643", type = "label")
```


### Multiple units, single variable

We will use `get_data_by_variable()`. We specify our single variable as `varId` string argument. If no `unitParentId` is provided function will return all available units for given variable. Setting `unitParentId` will return all available children units (on all levels). To narrow units level set `unitLevel`. Optionally we can specify interested years of data. If not all available years.

```{r  }
get_data_by_variable("420", unitParentId = "011210000000")
get_data_by_variable("420", unitLevel = "4")
```


## Useful tools

The `bdl` package provide couple of additional functions for summarizing and visualizing data.


### Summary

Data downloaded via `get_data_by_unit()` and `get_data_by_variable()` van be easily summarized by `summary()`:
```{r  }
df <- get_data_by_variable(varId = "3643", unitParentId = "010000000000")
summary(df)
```


### Plotting

Plotting functions in this package are interfaces to data downloading functions. Some of them require specifying `data_type` - method for downloading data and rest of arguments will be relevant to specified `data_type` function. Check documentation for more details.

```{r fig.height=5, fig.width=10}
line_plot(data_type = "unit", unitId = "000000000000", varId = c("415","420"))
```
```{r fig.height=5, fig.width=10}
pie_plot(data_type ="variable" ,"60559", "2006", unitLevel = "2")
```

Scatter plot is unique - requires vector of only 2 variables.
```{r fig.height=5, fig.width=10}
scatter_2var_plot(data_type ="variable" ,c("60559","415"), unitLevel = "2")
```

### Map generation

With `bdl` package comes `bdl.maps` dataset containing spatial maps for each Poland's level. `generate_map()` use them to generate maps filled with bdl data. Use `unitLevel` to change type of map. The lower level is chosen the slower map will generate as it has more spatial data to process.
```{r fig.height=5, fig.width=8, eval=FALSE}
generate_map(varId = "60559", year = "2017", unitLevel = 3)
```

### Multi download

Downloading functions `get_data_by_unit()` and `get_data_by_variable()` have alternative "multi" downloading mode. Function that would work for example single unit, if provided a vector will make additional column with values for each unit provided:
```{r  }
get_data_by_unit(unitId = c("023200000000", "020800000000"), varId =  c("3643", "2137", "148190"))
```
Or multiple variables for `get_data_by_variable()`:
```{r  }
get_data_by_variable(varId =c("3643","420"), unitParentId = "010000000000")
```

This type of downloading is removing some of columns from original "single" data download.

This mode works for locality version as well.
